Cross-national studies have not sufficiently accounted for non-independence between nations

### 2022-03-14 manuscript.Rmd
add language family data file

---
title             : "Cross-national studies have not sufficiently accounted for non-independence between nations"
shorttitle        : "Non-independence between nations"

@@

abstract: |
 Cross-national analyses can illuminate patternings of national-level outcomes around the world. But since nations are related via spatial proximity and shared cultural ancestry, cross-national analyses often violate statistical assumptions of non-independence, inflating false positive rates. Here, we show that past work in economics and psychology has not sufficiently accounted for non-independence between nations. In a review of the 100 highest-cited cross-national studies of economic development and cultural values, we find that controls for non-independence are rare and have not increased in frequency over the past 30 years. When studies do attempt to control for non-independence, our simulations suggest that the methods used continue to produce false positives. In replications of a subsample of twelve cross-national relationships, we show that half fail to replicate when controlling for spatial and/or cultural non-independence using global proximity matrices. We urge social scientists to sufficiently control for non-independence when reporting cross-national associations.

@@

Given the importance of nations for structuring human behaviour, there is justifiably huge interest in statistical analyses that attempt to predict variation in national-level outcomes of all kinds. At the time of writing, a search in Google Scholar for the term "cross-national study" returned 180,000 hits. The standard practice for cross-national analyses is to conduct bivariate correlations or multiple regressions with individual data points representing different nations. Such analyses widen the scope of social science beyond Western populations[@Henrich2010;@Pollet2014] and have been used to study, among other topics, the causes of variation in the economic wealth of nations[@Benhabib1994;@Comin2010;@LaPorta1997;@Sachs2001], global patterings of cultural norms and values[@Fincher2008;@Gelfand2011;@Hofstede2001;@Inglehart2000], and the universality and diversity of human behaviour and psychology around the world[@Awad2018;@Rhoads2021;@Schulz2019;@Thomson2018].

However, cross-national analyses are complicated by the fact that nations are not statistically independent data points. Unlike independent random samples from a population, nations are related to one another in a number of ways. First, nations that are closer to one another tend to be more similar than distant nations. This phenomenon is known as spatial non-independence[@Kissling2008], and it occurs because nations in close spatial proximity share characteristics due to local cultural diffusion of ideas[@Hewlett2002] and common climates and environments[@Kissling2008]. For example, the neighbouring African nations Zambia and Tanzania have similar levels of terrain ruggedness, which has been used to partially explain their similar levels of economic development[@Nunn2012]. This pattern conforms to Tobler's first law of geography: "everything is related to everything else, but near things are more related than distant things"[@Tobler1970] (p. 236).

Second, nations with shared cultural ancestry tend to be more similar than culturally unrelated nations. This is known as cultural phylogenetic non-independence[@Tylor1889], and occurs because related nations share cultural traits inherited via descent from a common ancestor. Shared cultural ancestry can result in a form of pseudoreplication, whereby multiple instances of the same trait across nations are merely duplicates of the ancestral original. For example, the related island nations New Zealand and Samoa express similar languages and customs due to cultural inheritance from a shared ancestor deep in the evolutionary history of Polynesian societies. More recently, shared ancestry is also crucial in explaining cultural similarities between colonial settlements and their colonisers, even those separated by large geographic distances (e.g. Spain and Argentina).

Spatial and cultural phylogenetic non-independence between nations make cross-national inference challenging. A fundamental assumption of regression analysis is that model residuals should be independently and identically distributed[@Jarque1987]. But without accounting for spatial or cultural non-independence between nations, model residuals can show structure that remains unaccounted for, violating this assumption (Figure \@ref(fig:spatialExample)). Treating nations as independent can thus inflate false positive rates[@Legendre1993], producing spurious "direct" relationships between variables that in fact only indirectly covary due to spatial or cultural connections (see Supplementary Figure \@ref(fig:dag) for an example causal model).

(ref:spatialExampleCaption) _An example of spatial non-independence in cross-national analysis._ National-level outcome and predictor variables for 25 nations are simulated with high levels of spatial non-independence. (top) Scatterplot visualising the data with fitted line from naive regression without accounting for spatial non-independence. Vertical line segments represent model residual errors, coloured by their sign and size. Countries are labelled with ISO 3166-1 alpha-2 letter codes. (bottom) Model residual errors plotted onto a world map. In this example, residuals clearly show spatial structure that remains unaccounted for.

```{r spatialExample, fig.width=6.5, fig.height=6, fig.cap="(ref:spatialExampleCaption)"}
tar_read(spatialExample)
```

@@

In order to motivate our research question, it is important to first show that economic development and cultural values are indeed spatially and culturally non-independent around the world. To this end, we used Bayesian multilevel models to simultaneously estimate geographic and cultural phylogenetic signal for the Human Development Index[@hdi] and two dimensions of cultural values from the World Values Survey, traditional vs. secular values and survival vs. self-expression values[@Inglehart2000]. For all three of these variables, we found that a substantial proportion of national-level variation was explained by spatial proximity and shared cultural ancestry between nations (Figure \@ref(fig:plotSignal)). Bayes Factors indicated strong evidence that these estimates of geographic and cultural phylogenetic signal differed from zero, aside from the geographic signal for survival values, for which the evidence was equivocal (see Supplementary Results). These findings emphasise the need to account for spatial and cultural non-independence in cross-national analyses of economic development and cultural values.

>>>>

---
title             : "Why non-independence matters for cross-national correlations"
shorttitle        : "Non-independence between nations"

@@

abstract: |
 Cross-national analyses can illuminate global patternings of national-level outcomes. But since nations are connected in various ways, such as via spatial proximity and shared cultural ancestry, cross-national analyses often violate assumptions of non-independence, inflating false positive rates. Here, we show that, despite being recognised as an important statistical pitfall for over 200 years, cross-national research in economics and psychology still does not sufficiently account for non-independence. In a review of the 100 highest-cited cross-national studies of economic development and values, we find that controls for non-independence are rare and have remained so over the past 30 years. When studies do include controls for non-independence, our simulations suggest that commonly used methods continue to produce false positives. In reanalyses of twelve cross-national relationships, we show that half are no longer significant after controlling for non-independence using global proximity matrices. We urge social scientists to sufficiently control for non-independence in cross-national research.

@@

Given the importance of nations for structuring human behaviour, there is justifiably huge interest in statistical analyses that attempt to predict variation in national-level outcomes of all kinds. At the time of writing, a search in Web of Science for the terms "cross-national", "cross-cultural", or "cross-country" in titles and abstracts returned approximately 80,000 unique hits. The standard practice for cross-national analyses is to conduct bivariate correlations or multiple regressions with individual data points representing different nations. Such analyses widen the scope of social science beyond Western populations[@Henrich2010;@Pollet2014] and have been used to study, among other topics, the causes of variation in the economic wealth of nations[@Benhabib1994;@Comin2010;@LaPorta1997;@Sachs2001], global patterings of cultural norms and values[@Fincher2008;@Gelfand2011;@Hofstede2001;@Inglehart2000], and the universality and diversity of human behaviour and psychology around the world[@Awad2018;@Rhoads2021;@Schulz2019;@Thomson2018].

However, cross-national analyses are complicated by the fact that nations are not statistically independent data points. Unlike independent random samples from a population, nations are related to one another in a number of ways. First, nations that are closer to one another tend to be more similar than distant nations. This phenomenon is known as spatial non-independence[@Kissling2008], and it occurs because nations in close spatial proximity share characteristics due to local cultural diffusion of ideas[@Hewlett2002] and regional variation in climate and environment[@Kissling2008]. For example, the neighbouring African nations Zambia and Tanzania have similar levels of terrain ruggedness, which has been used to partially explain their similar levels of economic development[@Nunn2012]. This pattern conforms to Tobler's first law of geography: "everything is related to everything else, but near things are more related than distant things"[@Tobler1970] (p. 236).

Second, nations with shared cultural ancestry tend to be more similar than culturally unrelated nations. This is known as cultural phylogenetic non-independence[@Tylor1889], and occurs because related nations share cultural traits inherited via descent from a common ancestor. Shared cultural ancestry can result in a form of pseudoreplication, whereby multiple instances of the same trait across nations are merely duplicates of the ancestral original. For example, despite being separated by over 1500 km of ocean, the related island nations Tonga and Tuvalu express similar languages and customs due to cultural inheritance from a shared ancestor deep in the evolutionary history of Polynesian societies. More recently, shared ancestry is also crucial in explaining cultural similarities between colonial settlements and their colonisers (e.g. Spain and Argentina).

Spatial and cultural phylogenetic non-independence between nations make cross-national inference challenging. A fundamental assumption of regression analysis is that model residuals should be independently and identically distributed[@Jarque1987]. But without accounting for spatial or cultural non-independence between nations, model residuals can show structure that remains unaccounted for, violating this assumption. Treating nations as independent can thus inflate false positive rates[@Legendre1993], producing spurious "direct" relationships between variables that in fact only indirectly covary due to spatial or cultural connections (see Supplementary Figure \@ref(fig:dag) for an example causal model).

@@

In order to motivate our research question, it is important to first show that economic development and cultural values are indeed spatially and culturally non-independent around the world. To this end, we used Bayesian multilevel models to simultaneously estimate geographic and cultural phylogenetic signal for the Human Development Index[@hdi] and two dimensions of cultural values from the World Values Survey, traditional vs. secular values and survival vs. self-expression values[@Inglehart2000]. For all three of these variables, we found that a substantial proportion of national-level variation was explained by spatial proximity and shared cultural ancestry between nations (Figure \@ref(fig:plotSignal)). In our models, shared cultural ancestry explained over half of the national-level variation in economic development and cultural values, with spatial proximity explaining most of the remaining national-level variance. Bayes Factors indicated strong evidence that these estimates of geographic and cultural phylogenetic signal differed from zero, aside from the geographic signal for survival values, for which the evidence was equivocal (see Supplementary Results). These findings emphasise the need to account for spatial and cultural non-independence in cross-national analyses of economic development and cultural values.
